---
title: "DeepPatientLevelPrediction Installation Guide"
author: "Egill Fridgeirsson"
date: '`r Sys.Date()`'
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead{}
  - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r  utils::packageVersion("PatientLevelPrediction")`}
  - \fancyfoot[CO,CE]{DeepPatientLevelPrediction Package Version `r  utils::packageVersion("DeepPatientLevelPrediction")`}
  - \fancyfoot[LE,RO]{\thepage}
  - \renewcommand{\headrulewidth}{0.4pt}
  - \renewcommand{\footrulewidth}{0.4pt}
output:
  html_document:
    number_sections: yes
    toc: yes
---

```{=html}
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Installing DeepPLP}
-->
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This vignette describes how to install the Observational Health Data Sciences
and Informatics (OHDSI) DeepPatientLevelPrediction package on Windows, macOS,
and Linux.

# Software Prerequisites

## Windows Users

Under Windows, install:

-   R (<https://cran.r-project.org/>) - R \>= 4.1.0; the latest release is recommended
-   Python - Python 3.14 is recommended; Python \>= 3.10 is supported
-   An R development environment, if desired
-   Java when using a JDBC-based database connection
-   [Rtools](https://cran.r-project.org/bin/windows/Rtools/) when installing packages from source

## Mac/Linux Users

Under macOS and Linux, install:

-   R (<https://cran.r-project.org/>) - R \>= 4.1.0; the latest release is recommended
-   Python - Python 3.14 is recommended; Python \>= 3.10 is supported
-   An R development environment, if desired
-   Java when using a JDBC-based database connection
-   Xcode command line tools on macOS when installing packages from source (`xcode-select --install`)

# Installing the Package

Install the released package from CRAN:

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy=FALSE, eval=FALSE}
install.packages("DeepPatientLevelPrediction")
```

The development version can instead be installed from GitHub. Development
versions may contain changes that have not yet been released.

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy=FALSE, eval=FALSE}
install.packages("remotes")
remotes::install_github("OHDSI/DeepPatientLevelPrediction@develop")
```

## Python environment

Model training and inference use PyTorch through `reticulate`. Python is not
required to install or load the R package. When Python functionality is first
used, `reticulate` can create a managed environment from the package
requirements on systems with internet access.

You can verify the active interpreter with:

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
library(DeepPatientLevelPrediction)
reticulate::py_config()
```

Advanced users, users with strict reproducibility requirements, or users in
air-gapped environments can manage the Python environment themselves and tell
`reticulate` which interpreter to use. One option is to create the environment
with `uv` and Python 3.14:

```bash
uv python install 3.14
uv venv --python 3.14
uv pip install polars tqdm pyarrow duckdb nvidia-ml-py numpy
uv pip install "torch==2.12.1" --index https://download.pytorch.org/whl/cpu/
```

The second `uv pip install` command installs the CPU build of PyTorch. If you want to train on a GPU, install the PyTorch build that matches your CUDA setup instead.

To force `reticulate` to use a manually managed interpreter, set `RETICULATE_PYTHON` in `.Renviron`.

For Linux/macOS:

```
RETICULATE_PYTHON="/path/to/project/.venv/bin/python"
```

For Windows:

```
RETICULATE_PYTHON="C:/path/to/project/.venv/Scripts/python.exe"
```

Then restart your R session.

Python 3.9 is end-of-life and should not be used. Python 3.10 is still supported, but Python 3.14 is recommended.

Accessing the `torch` helper or starting model training triggers `reticulate`
to resolve the Python requirements if `RETICULATE_PYTHON` is not configured.

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy=FALSE, eval=FALSE}
library(DeepPatientLevelPrediction)
torch$randn(10L)
```

This should print out a tensor with ten different values.

On Windows, close other R sessions that are using
`DeepPatientLevelPrediction` or its dependencies before updating the package;
open sessions can lock installed files.

# Testing Installation

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
library(DeepPatientLevelPrediction)

torch$randn(10L)
```

The R-side model settings can be created without initializing Python:

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy=FALSE}
modelSettings <- DeepPatientLevelPrediction::setResNet(
  numLayers = 2L,
  sizeHidden = 64L,
  hiddenFactor = 1L,
  residualDropout = 0,
  hiddenDropout = 0.2,
  sizeEmbedding = 64L,
  estimatorSettings = DeepPatientLevelPrediction::setEstimator(
    learningRate = 3e-4,
    weightDecay = 1e-6,
    device = "cpu",
    batchSize = 128L,
    epochs = 3L,
    seed = 42L
  ),
  hyperParamSearch = "random",
  randomSample = 1L
)

stopifnot(inherits(modelSettings, "modelSettings"))
```

To run an end-to-end patient-level prediction example, continue with the
[first-model vignette](FirstModel.html).

# Acknowledgments

Considerable work has been dedicated to provide the `DeepPatientLevelPrediction` package.

```{r tidy=TRUE,eval=TRUE}
citation("DeepPatientLevelPrediction")
```

**Please reference this paper if you use the PLP Package in your work:**

Reps JM, Schuemie MJ, Suchard MA, Ryan PB, Rijnbeek PR. Design and implementation of a standardized framework to generate and evaluate patient-level prediction models using observational healthcare data. J Am Med Inform Assoc. 2018;25(8):969-975. doi:10.1093/jamia/ocy032.
